Fine-scale Data Reveals Neighborhood Inequalities Effects of Heat on Expressed Mood

### 2021-06-23 doc/temp-sentiment.tex
Edits from Overleaf

%https://www.overleaf.com/project/5f918edd1afc380001feca0d
\title{Fine-scale Twitter Data Reveal Neighborhood Inequalities in Mental Health Effects of High Temperatures}

@@

\begin{abstract}
%Limit 150 words: https://www.nature.com/nclimate/about/content
%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health.  Indeed, traditional analyses of heat and mental health outcomes using data collected by public agencies have found strong associations between elevated temperatures and outcomes such as suicides and mental-health related hospitalizations.  However, these studies, which typically use data reported at the city or county level, have found no difference in vulnerability based on income or race.  We overturn this finding and show that there are stark differences in vulnerability using a novel data source: expressed sentiment in a quarter of a billion geolocated tweets, matched with prevailing weather as well as neighborhood economic and demographic conditions.  We find that increased temperatures worsen expressed sentiment in all areas, but that this effect is much stronger in poor and Black neighborhoods.
\end{abstract}
\begin{document}

@@

Findings from large-scale studies of relatively uniform vulnerability to heat are surprising given minority groups and poorer people are more likely to be exposed to undesirable temperatures and otherwise less able to mitigate the effects of higher heat. For example, people from poorer socio-economic backgrounds are more likely to work outside rather than indoors \cite{Gubernot2014Oct}, more likely to rely on public transportation, bicycling, or walking instead of commuting in their own air-conditioned car \cite{Karner2015Dec} Also, while minority groups and poorer people could be expanding their livelihood activities to increase their income levels by engaging in a range of activities, exclusions based on factors like race and gender limit such opportunities (Hallegatte et al. 2015). Diverse dimensions of the factors of inequalities have implications on coping abilities and further aggravate vulnerability. Thus, particular groups and conditions, for example race/ethnicity and density (urbanization) have been identified as having differential exposure or vulnerability to extreme events (Cardona et al 2012). For instance, there is evidence for increased vulnerability to extreme events among migrant groups because of language barriers hence inability to understand extreme event-related information and also their prioritization of finding employment and housing (Thomas et al 2019). Heterogeneities in vulnerability are also more pronounced within city regions rather than between cities. In the United States, considerable regional variability is evident (Cardona et al 2012). Some studies find large differences in vulnerability between neighborhoods for a variety of impacts of heat on physical health  \cite{Belanger2015Mar, Uejio2011Mar} and mental health for other types of climate shocks like hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}. So there are strong reasons to expect heterogeneities in the impact of heat on mental health outcomes. with these . Importantly, different degrees of inequality combine and interact for systems that relate directly to differential levels of social vulnerability, both in normal times and in the context of a disaster (Kuran et al 2020). This refers to the ‘concept of intersectionality’, where according to Kuran et al 2020, a factor may be a significant variable but only when allied with other variables. It primarily concerns how intersecting factors be it socioeconomic factors affect individuals who face multiple social inequities, with consequent multiple marginalisation.

While public health data on mental health outcomes is not available at the neighborhood scale, spontaneous, in situ data from the social media platform Twitter can provide an important, inexpensive and timely indicator of mental health at extremely precise spatial and temporal scales. Many studies use the mood expressed in social media as an indicator of mental health \cite{Edo-Osagie2020Jul, Sinnenberg2016Dec}. For example, studies have used Twitter data to identify the onset of Post Traumatic Stress Disorder (PTSD) in individuals even before their formal diagnosis \cite{Reece2017Oct}. Similar methods applied to posts on Facebook can predict the onset of depression \cite{Eichstaedt2018Oct}. In London, day-to-day changes in mental health indicators derived from the text of tweets were associated with changes in mental health crisis episodes \cite{Kolliakou2020Feb}. Across the USA, more positive tweets are associated with a variety of metrics of human well-being \cite{Mitchell2013May}. Finally, the mood expressed in tweets has been strongly associated with local weather conditions \cite{baylis_weather_2018, hannak_tweetin_2012} and been used as evidence to support the temperature-suicide relationship \cite{Burke2018Aug}. Thus, posts on Twitter provide an indicator of mental health at fine enough spatial scales to examine neighborhood effects on vulnerability to higher temperatures. We therefore draw on a data set of 243 million geolocated tweets from across the continental United States and covering the period 2009-2019, each paired local weather and neighborhood conditions, to explore how neighborhood conditions affect the relationship between heat and mental heath.

@@

\subsection*{Overall Effect}
We examine the relationship between the sentiment expressed in tweets and the prevailing Wet Bulb Globe Temperature (WBGT), a temperature metric that accounts for dry-bulb temperature, humidity, wind speed, and solar radiation to more accurately describe the effects of heat on the human body \cite{budd2008wet}. Controlling for a variety of fixed effects across all tweets, we find higher temperatures are associated with lower sentiment (i.e., worsening mood; See Fig. \ref{fig:wbgt}).  Sentiment is highest at 5\textdegree C WBGT (a dry bulb temperature typically around 12\textdegree C/54\textdegree F) with the largest declines in sentiment between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F). Sentiment also declines with colder temperatures (below 5\textdegree C WBGT), although only slightly. We ran models using multiple measures of sentiment and find similar results and patterns across all sentiment measures (see Supplement).

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{../res/wbgt.png}
  \caption{Relationship between Wet Bulb Globe Temperature (WBGT) and sentiment from text in tweets.  As temperatures increase above 5\textdegree C WBGT, sentiment rapidly declines.}
  \label{fig:wbgt}
\end{figure}

@@

We explored the relationship between neighborhood racial group characteristics and temperature and sentiment and find the effects of heat are felt disproportionately in neighborhoods that are majority Black (see Fig \ref{fig:hetero}). Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the sentiment of tweets in majority Black neighborhoods decrease four times as much as the sentiment of people in other neighborhoods. Additionally, at modest to moderate temperatures of 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority Hispanic neighborhoods have slightly lower sentiment than people in majority white or other neighborhoods, although this gap narrows at higher temperatures.

Exploring the combined effects of neighborhood percent minority race and income in moderating the impacts of heat on mental health, we find separate effects for these two variables (see Supplement). In other words, neither neither percent minority population nor income alone account for the heterogeneity in vulnerability as neighborhoods that are both poorer and higher percent Black are more affected by heat than than neighborhoods that are just poorer or higher proportion Black.

\begin{figure}[H]
\centering
  \includegraphics[width=\linewidth]{../res/wbgt_combined.png}
    \caption{Effect of changes in wet-bulb globe temperature on expressed sentiment relationship as moderated by neighborhood income (a) and race (b).}
\label{fig:hetero}
\end{figure}

@@

\subsection{Temperature Effects by Time of Day}
In addition to providing data with high spatial resolution, twitter data also comes with very high temporal resolution as each tweet is time-stamped.  Thus, we were able to examine the impact of heat on sentiment over the course of the day (See Fig. \ref{fig:ts-wbgt}).  We found that heat is associated with improved sentiment for a brief period in the late morning, and then has a consistently negative effect on sentiment during the latter half of the day, from noon until 9pm.  This effect weakens in the early evening through midnight, and then increases greatly throughout the night.  Heat has the greatest effect on expressed sentiment at 6am.

@@

\subsection*{Temperature Effects by Combined Statistical Area}
We examine the impact of temperature on sentiment across Combined Statistical Areas (CSAs) and Metropolitan Statistical Areas (MSAs) containing over 1 million people (hereafter: city regions). We find that higher temperatures are associated with declining sentiment in 83.6\% of city regions (see Fig. \ref{fig:map}) and the effect is significant in 35.3\% of these. Furthermore, we find no correlation between higher temperatures and increased sentiment in any of the city regions.

Increasing exposure from lower to higher temperatures - 5\textdegree C WBGT (12\textdegree C/54\textdegree F) to 30\textdegree C WBGT (42\textdegree C/108\textdegree F) - increases disparities in sentiment across income groups for 68.9\% of city regions, with 35.7\% of these having a statistically significant effect. City regions with a significantly unequal effect were located across the USA although they were most common in the mid-Atlantic region. Additionally, many city regions in the southwest had an inverted effect, where higher temperatures actually narrowed the gap in sentiment between wealthier and poorer neighborhoods, including Denver and Oklahoma City areas, which had a statistically significant inverted effect.

Given the strong effect of temperature on sentiment in majority Black neighborhoods compared to other non-majority Black neighborhoods, we also examined increases in sentiment gaps between neighborhoods in the 95th percentile for Black population (77.2\% Black) and neighborhoods in the 5th percentile for Black population (0\% Black). We excluded city regions with a Black population less than 5\% of the total population. We found increasing inequalities in sentiment for more Black neighborhoods in 64.7\% of city regions, with a significant effect in 21.2\% of those (7 city regions).  Additionally, three sunbelt city regions had a significant inverted effect. The patterns of inequalities in the impacts of heat on sentiment by race were similar to inequalities by income: city regions with large and significant inequalities were located in the mid-Atlantic and Midwest.

\begin{figure}[H]
\centering
  \includegraphics[width=0.5\linewidth]{../res/map_combined.png}
  \label{fig:map}
    \caption{Figure (a) shows the change in sentiment for a 25-degree increase in WBGT for city regions with a population of over 1 million.  Figures (b) and (c) show inequalities in impacts of temperature on sentiment for city regions with over 1 million people.  The value shown is the predicted change in the size of the gap in VADER score as temperatures increase from 5\textdegree C WBGT to 30\textdegree C WBGT between the 5th and 95 percentile for income (b) and the 95th and 5th percentile for percent Black (c). City regions with a statistically significant effect ($p < 0.05$) are outlined in Black.  For racial inequalities (c), only cities where the proportion of Black people is $> 5\%$ of the population are included.}
\end{figure}

@@

Our study found disproportionate effects of climate change on identifiable and marginalized populations. We however found that majority Black neighborhoods are much more affected by higher temperatures than neighborhoods with a non-Black majority. This is surprising, given that Hispanics can be marginalized in income, language and housing. This finding may be because the racial category of Hispanic is more broad and encompasses many more groups with more diverse histories and income levels than Black Americans. It may also be due to that fact that more marginalized and heat-vulnerable Hispanic people were more likely to tweet in the Spanish language, which we did not include our analysis.  Additionally, there are many other marginalized groups in America, particularly indigenous people, that we did not have sufficient data to examine with respect to heat and mental health.  Finally, while race is correlated with income in the US, we found that race contributed to vulnerability to heat, independently of income. This suggests that racism and patterns of housing discrimination like the legacy of redlining contribute to making Black Americans more vulnerable to heat, event at similar levels of income. Thus, our study shows that, the degree of vulnerability could be related to a category of different dimensions (like combinations of race and housing) and not to a single dimension (race or income). This explains how complex vulnerability can be assessed. It further concurs with the ‘concept of intersectionality’ where the degree of vulnerability does not depend on a single dimensional attribution but results from a complex relationship between different factors such as a combination of race, housing and income (Kuran et al 2020).

@@

% Caveats
% Sentiment is a fuzzy indicator but we have BIG DATA
While these findings are robust to different model specifications and sentiment metrics, there are some caveats and limitations. One limitation of data from Twitter is that a tweet is not a precise measurement of a discrete mental health outcome and only provides a rough indicator a user's mental state based on the vocabulary in the tweet.  Moreover, while sentiment analysis algorithms have gotten increasingly sophisticated at estimating the mood in a body of text, a user's expressed sentiment in a tweet is highly variable and is mostly affected by factors like current events and the user's personal life, with local weather conditions having only a small impact.  We overcome this limitation by using an extremely large data set of 243 million tweets, allowing us to control for a wide variety of spatial and temporal effects to estimate population-level changes in sentiment in relation to the weather.

One caveat associated with this data is that we were only able to locate the tweets within the census block where the tweet was sent - we did not to infer where the person sending the tweets typically lived or where they had been.  Many low-income and Black people commute during the day to work in the service sector in higher-income areas, so these results may in fact under-estimate the impacts of higher temperatures on mental health for poor and minority individuals.  Moreover, in some cases the income level of a census block may be only weakly indicative of the wealth of the people who are typically found that census block.  For example, some public spaces are estimated to have very low income levels even though people from a variety of income levels may occupy those spaces throughout the day.  Additionally, neighborhoods of young college students are also typically estimated to have low income levels, even through college students are wealthier than the average American.  Again, these issues mean we are likely under-estimating the true effect of heat on mental health for poor and Black people.  A final issue with using Twitter data is that, while Twitter is used by more than one in five Americans, Twitter users may not be representative of the general population, as they are typically younger, wealthier, and more educated \cite{Pew2020Sep}.  Nevertheless, we found a large volume of tweets across all neighborhood types.

@@

\section*{Methods}
\subsection*{Tweets \& Sentiment}
We used data from 243 million geo-located English-language tweets from the continental United States from the years 2009 to 2019.  The Twitter data consists of publicly posted messages, or tweets, that are short status messages users post to the platform.  We only considered users’ original content, thus did not include retweets in our analysis.  Additionally, we excluded all tweets that contained weather-related terms, to ensure that the sentiment expressed in the tweets was reflective of a user's mental state, and not a commentary on the weather.

@@

The VADER method yields a value for the mood of a tweet, with a score of 0 for neutral tweets, a score $> 0$ for positive-mood tweets and a score $< 0$ for negative-mood tweets.  In addition to the VADER method used in the body of this paper, we also conduct our analysis using the Hedonometer and XXXXX sentiment analysis methods, with similar results (See Supplement).

@@

\subsection*{Socio-Econoimc Data}
We used data from the American Community Survey (ACS) administered by the US Census to estimate income levels and the racial composition of neighborhoods where tweets were located.  Data was at the level of the census block group, the smallest unit for which the US Census releases public data. Following Census categories, we report racial characteristics of neighborhoods as majority population of four broad racial categories - non-Hispanic Black, non-Hispanic white, Hispanic of any race, and other, which includes Native American, multi-racial, Asian-American and Pacific Islander.

ACS data is released to cover five-year periods.  We therefore matched each tweet with census block group data from the year at the middle year of each survey's five-year range.  For example, tweets from 2014 were matched to data from the 2012-2016 ACS.  Because the most recent available ACS was from 2014-2018, all tweets from year years greater than 2016 were matched to this dataset.  Data was downloaded from the IPUMS NHGIS service provided by the University of Minnesota.

@@

In this paper, we give income and racial percentiles based on the observed percentiles across all tweets in our data set.

\subsection*{Modeling}
We assessed how expressed sentiment is affected by Wet Bulb Globe Temperature using segmented regressions, controlling for precipitation, and shortwave solar radiation (sunshine), as well as the following spatio-temporal fixed effects: the day of the week, the time of day, the day of the year, the year, the month, and the county.

@@

Where $y$ is the sentiment of a tweet, $t$ is the wet bulb globe temperature at the hour of the tweet, $p$ is a binary variable indicating whether it rained at the hour of the tweet, $s$ is the income shortwave radiation, or sunshine, in $W/m^2$, at the hour of the tweet, $\Phi$ is the spatio-temporal fixed effects, $\epsilon$ is the normally-distributed errors, and $f_t$ represented the segmented regression for $t$, with knots at -10\textdegree, 0\textdegree, 5\textdegree, 10\textdegree, 15\textdegree, 20\textdegree, and 25\textdegree C WBGT.  We estimate the 95\% confidence interval of our models using 80-fold bootstrapping.

@@

For our analyses by city, we run models with a similar form to Model \ref{mod:2}, including the same fixed effects, although we fit a linear effect for $f_t()$.  Additionally, for exploring the vulnerability of Black americans, we use a continuous variable for $m$ indicating the percentage of a neighborhoods population that is Black.  Finally, because we are not accounting for non-linearities in $f_t()$, we exclude tweets with a WBGT of less than 5\textdegree C, as our initial analysis showed that this is the temperature at which sentiment peaks.

@@

\section*{Author Contributions}
Author contributions: M.W.C., J.O.-G., A.S. and P.B. designed the research and modeling strategy; A.S. provided data; M.W.C and Z.L. prepared data; M.W.C. and J.O.-G. analyzed data; and M.W.C., J.O.-G., J.L., and P.A.W. wrote the paper.

\noindent This paper was the result of a SESYNC graduate pursuits project co-lead by M.W.C. and J.O.-G.

>>>>

%https://www.overleaf.com/project/5f918edd1afc380001feca0d
\title{Fine-scale Data Reveals Neighborhood Inequalities Effects of Heat on Expressed Mood}

@@

\begin{abstract}
%Limit 150 words: https://www.nature.com/nclimate/about/content
%Actually there are lots of examples of articles with > 150 words, but I think we should keep it concise
Higher temperatures associated with climate change are expected to have major impacts on human mental health.  Indeed, traditional analyses of heat and mental health outcomes using data collected by public agencies have found strong associations between elevated temperatures and outcomes such as suicides and mental-health related hospitalizations.  However, these studies, which typically use data reported at the city or county level, have found no difference in vulnerability based on income or race.  We overturn this finding and show that there are stark differences in vulnerability using a novel data source: expressed sentiment in a quarter of a billion geolocated Tweets, matched with prevailing weather as well as neighborhood economic and demographic conditions.  We find that increased temperatures worsen expressed sentiment in all areas, but that this effect is much stronger in poor and Black neighborhoods.
\end{abstract}
\begin{document}

@@

Findings from large-scale studies of relatively uniform vulnerability to heat are surprising given minority groups and poorer people are more likely to be exposed to undesirable temperatures and otherwise less able to mitigate the effects of higher heat. For example, people from poorer socio-economic backgrounds are more likely to work outside rather than indoors \cite{Gubernot2014Oct}, more likely to rely on public transportation, bicycling, or walking instead of commuting in their own air-conditioned car \cite{Karner2015Dec}. Minority groups and poorer peoples frequently face other systemic and societal exclusions that can limit opportunities for advancement which can limit coping abilities and further aggravate and compound vulnerabilities \cite{hallegatte2016shock}, seomtimes referred to as 'intersectionality' \cite{Kuran2020Nov}. Thus, particular groups and conditions like race/ethnicity and density (urbanization) can lead to compounding and differential exposures or vulnerabilities of these groups to extreme events \cite{Cardona2012Jan}. For instance, migrant groups are sometimes especially vulnerable to heat waves because language barriers make it difficult or impractical to interpret public safety information, compounded during heat-related events if these groups are prioritize finding employment and housing \cite{Thomas2019Mar}. Heterogeneities in vulnerability are also more pronounced within city regions rather than between cities, particularly in the United States where there may be considerable regional variability \cite{Cardona2012Jan}. Some studies find large differences in vulnerability between neighborhoods for a variety of impacts of heat on physical \cite{Belanger2015Mar, Uejio2011Mar} and mental health for other types of climate shocks like hurricanes \cite{ferre2019hurricane, Gruebner2015Jun}. Importantly, different degrees of inequality combine and interact for systems that relate directly to differential levels of social vulnerability, both in normal times and in the context of a disaster \cite{Kuran2020Nov}. So there are strong reasons to expect heterogeneities in the impact of heat on mental health outcomes. It primarily concerns how intersecting factors be it socioeconomic factors affect individuals who face multiple social inequities, with consequent multiple marginalisation.

While public health data on mental health outcomes is not available at the neighborhood scale, spontaneous, in situ data from the social media platform Twitter can provide an important, inexpensive and timely indicator of mental health at extremely precise spatial and temporal scales. Many studies use the mood expressed in social media as an indicator of mental health \cite{Edo-Osagie2020Jul, Sinnenberg2016Dec}. For example, studies have used Twitter data to identify the onset of Post Traumatic Stress Disorder (PTSD) in individuals even before their formal diagnosis \cite{Reece2017Oct}. Similar methods applied to posts on Facebook can predict the onset of depression \cite{Eichstaedt2018Oct}. In London, day-to-day changes in mental health indicators derived from the text of Tweets were associated with changes in mental health crisis episodes \cite{Kolliakou2020Feb}. Across the USA, more positive Tweets are associated with a variety of metrics of human well-being \cite{Mitchell2013May}. Finally, the mood expressed in Tweets has been strongly associated with local weather conditions \cite{baylis_weather_2018, hannak_tweetin_2012} and been used as evidence to support the temperature-suicide relationship \cite{Burke2018Aug}. Thus, posts on Twitter provide an indicator of mental health at fine enough spatial scales to examine neighborhood effects on vulnerability to higher temperatures. We therefore draw on a data set of 243 million geolocated Tweets from across the continental United States and covering the period 2009-2019, each paired local weather and neighborhood conditions, to explore how neighborhood conditions affect the relationship between heat and mental heath.

@@

\subsection*{Overall Effect}
We examine the relationship between the sentiment expressed in Tweets and the prevailing Wet Bulb Globe Temperature (WBGT), a temperature metric that accounts for dry-bulb temperature, humidity, wind speed, and solar radiation to more accurately describe the effects of heat on the human body \cite{budd2008wet}. Controlling for a variety of fixed effects across all Tweets, we find higher temperatures are associated with lower sentiment (i.e., worsening mood; See Fig. \ref{fig:wbgt}).  Sentiment is highest at 5\textdegree C WBGT (a dry bulb temperature typically around 12\textdegree C/54\textdegree F) with the largest declines in sentiment between 20\textdegree -25\textdegree C WBGT (dry bulb 29\textdegree-36\textdegree C/84\textdegree-97\textdegree F). Sentiment also declines with colder temperatures (below 5\textdegree C WBGT), although only slightly. We ran models using multiple measures of sentiment and find similar results and patterns across all sentiment measures (see Supplement).

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{../res/wbgt.png}
  \caption{Relationship between Wet Bulb Globe Temperature (WBGT) and sentiment from text in Tweets.  As temperatures increase above 5\textdegree C WBGT, sentiment rapidly declines.}
  \label{fig:wbgt}
\end{figure}

@@

We explored the relationship between neighborhood racial group characteristics and temperature and sentiment and find the effects of heat are felt disproportionately in neighborhoods that are majority Black (see Fig \ref{fig:hetero}). Relative to an optimum temperature of 5\textdegree C WBGT (12\textdegree C/54\textdegree F), as temperatures increase to 30\textdegree C WBGT (42\textdegree C/108\textdegree F), the sentiment of Tweets in majority Black neighborhoods decrease four times as much as the sentiment of people in other neighborhoods. Additionally, at modest to moderate temperatures of 10\textdegree C WBGT (19\textdegree C/66\textdegree F) to 25\textdegree C WBGT (36\textdegree C/97\textdegree F), people in majority Hispanic neighborhoods have slightly lower sentiment than people in majority white or other neighborhoods, although this gap narrows at higher temperatures.

\begin{figure}[H]
\centering
  \includegraphics[width=\linewidth]{../res/wbgt_combined.png}
    \caption{Effect of changes in WBGT on expressed sentiment as moderated by neighborhood income (a) and race (b).}
\label{fig:hetero}
\end{figure}

Finally, given that both low-income and majority black neighborhoods are more vulnerable to heat waves, we model the combined effects of race and sentiment together to explore the intersection of race and income.  We model how the relationship between temperature and sentiment is moderated by the per-capita income of a neighborhood as well as the percentage of a neighborhood's inhabitants that are black.  Additionally, we compare AICs for models with no term for vulnerability, a race term for vulnerability, an income term for vulnerability, and both race and income terms for vulnerability.

\begin{minipage}{.65\textwidth}
\centering
\includegraphics[width=\textwidth]{../res/wbgt-income-race.png}
\captionof{figure}{Relationship between WBGT and sentiment from text in Tweets, moderated by the interaction of race and income.}
\label{fig:race-income}
\end{minipage}\hfill
\begin{minipage}{.35\textwidth}
\centering
\begin{tabular}{| l | l |}
\hline
Terms & AIC \\
\hline
Neither & 277270561 \\
Income  & 277182239 \\
Race    & 277143579 \\
Both    & 277108676 \\
\hline
\end{tabular}
\label{tab:aics}
\captionof{table}{Akaike information criterion (AIC) of models with interaction terms for race, income, both race and income, as well as neither race nor income.  The model with the best fit by AIC includes both race and income interaction terms.}
\end{minipage}

Graphing the combined effects of race and income, we find that race has a larger impact on vulnerability than income (see Fig. \ref{fig:race-income}).  A fully black neighborhood at any income level will be much more impacted by heat than a poor neighborhood with few black inhabitants.  Additionally, neighborhoods that are largely black are also more affected by cold temperatures, although there is no clear effect of income moderating the relationship between cold temperatures and sentiment.  Using AIC as an indicator of how different terms affect model quality (see Tab. \ref{tab:aics}), we find that a model including both terms performs the best, indicating that both race and income matter for the impact of heat on sentiment.  Moreover, including only a race term improves the model more than including only an income term, further supporting the conclusion that race is more important than income for explaining the vulnerability of sentiment to heat.

@@

\subsection{Temperature Effects by Time of Day}
In addition to providing data with high spatial resolution, Twitter data also comes with very high temporal resolution as each tweet is time-stamped.  Thus, we were able to examine the impact of heat on sentiment over the course of the day (See Fig. \ref{fig:ts-wbgt}).  We found that heat is associated with improved sentiment for a brief period in the late morning, and then has a consistently negative effect on sentiment during the latter half of the day, from noon until 9pm.  This effect weakens in the early evening through midnight, and then increases greatly throughout the night.  Heat has the greatest effect on expressed sentiment at 6am.

@@

@@

Our study found disproportionate effects of climate change on identifiable and marginalized populations. We however found that majority Black neighborhoods are much more affected by higher temperatures than neighborhoods with a non-Black majority. This is surprising, given that Hispanics can be marginalized in income, language and housing. This finding may be because the racial category of Hispanic is more broad and encompasses many more groups with more diverse histories and income levels than Black Americans. It may also be due to that fact that more marginalized and heat-vulnerable Hispanic people were more likely to tweet in the Spanish language, which we did not include our analysis.  Additionally, there are many other marginalized groups in America, particularly indigenous people, that we did not have sufficient data to examine with respect to heat and mental health.  Finally, while race is correlated with income in the US, we found that race contributed to vulnerability to heat, independently of income. This suggests that racism and patterns of housing discrimination like the legacy of redlining contribute to making Black Americans more vulnerable to heat, event at similar levels of income. Thus, our study shows that, the degree of vulnerability could be related to a category of different dimensions (like combinations of race and housing) and not to a single dimension (race or income). This explains how complex vulnerability can be assessed. It further concurs with the ‘concept of intersectionality’ where the degree of vulnerability does not depend on a single dimensional attribution but results from a complex relationship between different factors such as a combination of race, housing and income \cite{Kuran2020Nov}.

@@

% Caveats
% Sentiment is a fuzzy indicator but we have BIG DATA
While these findings are robust to different model specifications and sentiment metrics, there are some caveats and limitations. One limitation of data from Twitter is that a tweet is not a precise measurement of a discrete mental health outcome and only provides a rough indicator a user's mental state based on the vocabulary in the tweet.  Moreover, while sentiment analysis algorithms have gotten increasingly sophisticated at estimating the mood in a body of text, a user's expressed sentiment in a tweet is highly variable and is mostly affected by factors like current events and the user's personal life, with local weather conditions having only a small impact.  We overcome this limitation by using an extremely large data set of 243 million Tweets, allowing us to control for a wide variety of spatial and temporal effects to estimate population-level changes in sentiment in relation to the weather.

One caveat associated with this data is that we were only able to locate the Tweets within the census block where the tweet was sent - we did not to infer where the person sending the Tweets typically lived or where they had been.  Many low-income and Black people commute during the day to work in the service sector in higher-income areas, so these results may in fact under-estimate the impacts of higher temperatures on mental health for poor and minority individuals.  Moreover, in some cases the income level of a census block may be only weakly indicative of the wealth of the people who are typically found that census block.  For example, some public spaces are estimated to have very low income levels even though people from a variety of income levels may occupy those spaces throughout the day.  Additionally, neighborhoods of young college students are also typically estimated to have low income levels, even through college students are wealthier than the average American.  Again, these issues mean we are likely under-estimating the true effect of heat on mental health for poor and Black people.  A final issue with using Twitter data is that, while Twitter is used by more than one in five Americans, Twitter users may not be representative of the general population, as they are typically younger, wealthier, and more educated \cite{Pew2020Sep}.  Nevertheless, we found a large volume of Tweets across all neighborhood types.

@@

\section*{Methods}
\subsection*{Tweets \& Sentiment}
We used data from 243 million geo-located English-language Tweets from the continental United States from the years 2009 to 2019.  The Twitter data consists of publicly posted messages, or Tweets, that are short status messages users post to the platform.  We only considered users’ original content, thus did not include reTweets in our analysis.  Additionally, we excluded all Tweets that contained weather-related terms, to ensure that the sentiment expressed in the Tweets was reflective of a user's mental state, and not a commentary on the weather.

@@

The VADER method yields a value for the mood of a tweet, with a score of 0 for neutral Tweets, a score $> 0$ for positive-mood Tweets and a score $< 0$ for negative-mood Tweets.  In addition to the VADER method used in the body of this paper, we also conduct our analysis using the Hedonometer and AFINN sentiment analysis methods, with similar results (See Supplement).

@@

\subsection*{Socio-Econoimc Data}
We used data from the American Community Survey (ACS) administered by the US Census to estimate income levels and the racial composition of neighborhoods where Tweets were located.  Data was at the level of the census block group, the smallest unit for which the US Census releases public data. Following Census categories, we report racial characteristics of neighborhoods as majority population of four broad racial categories - non-Hispanic Black, non-Hispanic white, Hispanic of any race, and other, which includes Native American, multi-racial, Asian-American and Pacific Islander.

ACS data is released to cover five-year periods.  We therefore matched each tweet with census block group data from the year at the middle year of each survey's five-year range.  For example, Tweets from 2014 were matched to data from the 2012-2016 ACS.  Because the most recent available ACS was from 2014-2018, all Tweets from year years greater than 2016 were matched to this dataset.  Data was downloaded from the IPUMS NHGIS service provided by the University of Minnesota.

@@

In this paper, we give income and racial percentiles based on the observed percentiles across all Tweets in our data set.

\subsection*{Modeling}
We assessed how expressed sentiment is affected by Wet Bulb Globe Temperature using segmented regressions, controlling for precipitation, shortwave solar radiation (sunshine), as well as the following spatio-temporal fixed effects: the day of the week, the time of day, the day of the year, the year, the month, and the county.

@@

Where $y$ is the sentiment of a tweet, $t$ is the wet bulb globe temperature at the hour of the tweet, $p$ is a binary variable indicating whether it rained at the hour of the tweet, $s$ is the income shortwave radiation, or sunshine, in $W/m^2$, at the hour of the tweet, $\Phi$ is the spatio-temporal fixed effects, $\epsilon$ is the normally-distributed errors, and $f_t$ represents the segmented effect for $t$, with knots at -10\textdegree, 0\textdegree, 5\textdegree, 10\textdegree, 15\textdegree, 20\textdegree, and 25\textdegree C WBGT.  We estimate the 95\% confidence interval of all our models using 80-fold bootstrapping.

@@

In addition to our models exploring how income and race moderate the effect of heat on sentiment individually, we fit an additional model with an interaction terms for both race and income.  Due to the complexity of a model with segmented terms interacting with two different variables, using a categorical variable for race would have yielded a model matrix too large for the largest cloud virtual machine available to us.  Thus, we used a continuous metric for both income as well as race (percent black).  This model extends our main model to the following form:

\begin{equation}
    y = \begin{tabular}{l}
    $\beta_0 + f_t(t) + f_{m_{i}t}(m_{i} t) + f_{m_{r}t}(m_{r} t) + 
    \beta_p p + \beta_{m_{r} p} m_{r} p + 
    \beta_{m_{i} p} m_{i} p + \beta_{m_{i} m_{r} p} m_{i} m_{r} p + $
    \\
    $\beta_s s + \beta_{m_{i} s} m_{i} s + 
    \beta_{m_{r} s} m_{r} s + \beta_{m_{i} m_{r} s} m_{i} m_{r} s + 
    \Phi + \epsilon$
    \end{tabular}
    \label{mod:race-inc}
\end{equation}

In this new model, $m_{i}$ is average income in the census block where the tweet originated, while $m_{r}$ is the racial composition of the census block.

Finally, we fit a model for how heat affects sentiment by time of day.  For this analysis, we used a varying-coefficient model, where the effect of heat is linear, but the coefficient for the effect varies non-linearly as a function of the time of day.  Because we modeled the effect of heat as linear, we excluded all tweets with an observed temperatures less than 5\textdegree C WBGT so that we were only including values at which our previous analyses suggested the heat-sentiment relationship is monotonic.  This final model took the following form:

\begin{equation}
    y = \beta_0 + f_h(h)t + \beta_p p + \beta_s s + \Phi + \epsilon
    \label{mod:tod}
\end{equation}

In this final model, $f_{h}$ is the coefficient for temperature, $t$, and varies non-linearly by time of day, $h$.  To ensure that the effect $f_{h}$ is cyclical, we estimated $f_{h}$ using 2nd-order B-splines, also known as a tent function basis \cite[Chapter~4.2]{wood2017generalized}, with knots every two hours.

@@

\section*{Author Contributions}
Author contributions: M.C., J.O.-G., A.S. and P.B. designed the research and modeling strategy; A.S. provided data; M.C and Z.L. prepared data; M.C. and J.O.-G. analyzed data; and M.C., J.O.-G., J.L., and P.A.W. wrote the paper.

\noindent This paper was the result of a SESYNC graduate pursuits project co-lead by M.C. and J.O.-G.
